Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2016-09-15 pkg/glmertree/paper/treatment_interactions.Rnw
Updated paper based on Niels comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@980 255f4249-2263-4417-b1ea-8498779e27a7

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are particularly useful for exploratory purposes, as they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions. Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.

In many instances, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. For tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection take into account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM tree.

The GLMM tree algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}), which offers a flexible framework for subgroup detection. For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}. GLMM tree allows for taking into account the clustered structure of datasets \cite<which is not done in other purely tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}, as well as treatment effect estimation with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we first introduce an artificial motivating dataset, which will be used to illustrate and explain treatment-subgroup interaction detection with GLM-based MOB and GLMM tree. In the Empirical Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects. In the Application, we apply the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. Finally, in the Discussion we summarize the results and point out some directions for future research.

@@

Participants were part of one of ten clusters, each with a different value for the intercept. Data was generated such that covariates and cluster-specific intercepts were uncorrelated. Also, 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, and 8\% of variance was due to cluster-specific variation.

@@

The rationale behind MOB is that a global parametric model may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parameters of more complex models across terminal nodes.

For example, let us take a global GLM to estimate the overall treatment effect in the artificial motivating dataset. The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function\footnote{An overview of notation used is provided in the appendix.}:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups. Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper. Also, we assume a continuous response variable, employing a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$. However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.

@@

The result of fitting a global GLM to the artificial motivating dataset is depicted in Figure~\ref{fig:fixedeffects}; the boxplots show the distribution of the posttreatment depression scores in both treatment groups. The global model does not describe the data well: there is substantial residual variance and the slope of the regression line is nearly zero. This does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.

The MOB algorithm can be used to detect such moderation, by testing for parameter stability over a set of auxiliary covariates or partitioning variables. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

@@

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ yields a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). The binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2). The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node or subgroup:
%
\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}
%
Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all coefficients interact with the subgroup indicator. Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{*}$ would have length $2J$, and contain the subgroup-specific fixed-effects coefficients.

@@

Figure~\ref{fig:example_mobtree} shows the GLM tree grown on the artificial motivating dataset. By using the three auxiliary covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the true treatment-subgroup interactions described above. However, the two subgroups in the middle do not represent true subgroups, which may be due to the clustered structure of the dataset not being taken into account.

@@

For estimating GLMM trees, we take this approach a step further: we take a GLM tree, instead of CART tree with constant fits to estimate the fixed-effects part of the GLMM. This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes. In addition, GLMM trees can be estimated for continuous, as well as binary and count variables. The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:

@@

In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the artificial motivating dataset is presented. As can be seen, by taking into account the clustering of observations, the true treatment subgroups have been recovered, and the spurious split involving the anxiety variable no longer appears in the tree.

@@

\section{Empirical Evaluation: Method}

To assess and compare the performance of GLMM tree, we performed three simulation studies. In Study I, we compared the accuracy of GLMM tree with that of GLM tree in datasets with treatment-subgroup interactions. In Study II, we compared the accuracy of GLMM tree with that of GLM tree in datasets without treatment-subgroup interactions, allowing for assessment of the Type-I error rate. In Study III, we compared the accuracy of GLMM tree with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.

@@

R \cite{R14} was used for generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions. For other response distributions, the \verb|glmtree| function would be available. For estimation of GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.

@@

@@

Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}) in each dataset. Consequently, predictive accuracy was only be assessed in datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions. To prevent overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, predictive accuracy was assessed using test datasets. Test datasets were generated from the same population as training datasets, but test observations were not drawn from the same clusters as the training observations, but from 'new' clusters.

\section{Empirical Evaluation: Results}

\subsection{Study I: Piecewise interactions}

For each cell of the data-generating design described above, 50 datasets were generated, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2$ = 32,400 training datasets with piecewise treatment-subgroup interactions. In these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure~\ref{fig:modelC}).

@@

Figure~\ref{fig:xyplot_treesize_interact} indicates that GLMM tree grows trees of about the true tree size in all conditions. On average, tree size increases slightly with sample size for GLMM tree, and dramatically for GLM tree. In the absence of random effects (i.e., $\sigma_{b}=0$), GLM and GLMM trees grow trees of about the same size. However, clear differences in tree size were observed when $\sigma_b > 0$. When sample size is small and random intercepts are not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average. When sample size is larger and random intercepts are not correlated with a partitioning covariate, GLM and GLMM trees are about equally sized. However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially wit larger sample sizes (i.e., 500 or 1000) and when $\sigma_b$ values (i.e., 10). This effect was somewhat stronger when the random intercept was correlated with a covariate that did not appear in the true tree.

@@

\subsubsection{Design}
For assessing Type-I error of GLM and GLMM tree, we generated datasets without treatment-subgroup interactions, in which there is only a main effect of treatment in the population. Put differently, there was only a single global value of $\beta_j = \beta$ in every dataset. For these datasets, we varied the same parameters of the data-generating process as in Study I. However, the sixth facet of the data-generating process had only two levels for these datasets, as there were no `true' partitioning variables in these datasets. Therefore, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2$ = 21,600 datasets without treatment-subgroup interactions were generated.

@@

\subsection{Study III: Linear and piecewise interactions}

\subsubsection{Design}
To compare the performance of GLMM tree with that of GLMMs with pre-specified interaction effects, we created datasets in which interactions were varied from purely linear to purely piecewise. The number of cells in the simulation design was reduced by limiting the fourth facet of the data-generating design to a single level ($M=25$ clusters). The fifth facet of the data-generating design was limited to two levels ($\sigma_b = 2.5$ and $\sigma_b = 7.5$). As in Study II, we limited the sixth facet to two levels (the random intercept was either correlated to one of the partitioning variables, or not). To counter unidentifiability due to potentially large numbers of pre-specified interaction terms in the GLMMs, we limited the number of covariates to $k=5$ and $k=10$ (instead of $k=5$ and $k=15$). Finally, to allow for a smooth transition from linear to piecewise interactions, the seventh facet of the interaction design (effect size of treatment differences) was replaced by a factor with three levels, controlling the type of interactions: purely linear interactions, purely piecewise interactions, and a combination of both. As a result, $50 \times 3 \times 2 \times 2 \times 1 \times 2 \times 2 \times 3$ = 7,200 training datasets were generated for this study.

All input variables were generated as described above. To generate purely linear interactions, outcome variable $y$ was generated as the sum of:

Note that the direction (positive or negative) and type (main or interaction) of these effects correspond to those in Figure~\ref{fig:modelC}. All interactions were created by using centered $U_k$ variables, calculated by subtracting the variable means. To generate purely piecewise interactions, we used the partition depicted in Figure~\ref{fig:modelC}. We calculated treatment-subgroup specific means $\hat{\mu}_{y}$ in a large dataset (N = $10^6$) with purely linear interactions. For each observation in datasets with purely piecewise interactions, this treatment-subgroup specific $\hat{\mu}_{y}$ value was imputed. To generate combined linear and piecewise interactions, $y$ values according to the purely piecewise and linear models were summed and divided by 2.

@@

<<linear-simulation-data>>=
load("pw_treespecs_long.dat")
treespecs.long$N <- factor(as.numeric(as.character(treespecs.long$N)), ordered=T)
treespecs.long$sigmab <- factor(as.numeric(as.character(treespecs.long$sigmab)), ordered=T)
treespecs.long$type <- factor(treespecs.long$type, ordered = TRUE, levels = 
c("linear", "both", "piecewise"))
@

@@

Overall, GLMM tree accurately recovered main and interaction effects in 53.58\% of datasets, wheareas GLMMs accurately recovered main and interaction effects in 34.81\% of datasets. To find the most important predictors of model accuracy, we performed an ANOVA, in which the algorithm type, sample size, the number of covariates and type of interaction (linear, piecewise or both) were found to be most important. These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_accuracy_pw}.

Figure~\ref{fig:xyplot_accuracy_pw} indicates a clear effect of sample size: when the number of covariates is large (i.e., 10) GLMMs do not recover all true effects in the dataset, whereas they always seem to recover the true effects when the number of covariates is small (i.e., 5), and the effects are partly linear. However, when the effects are purely piecewise, GLMMs never recover the true effects in the dataset, unless sample size is very large (i.e., $N = 1000$). At the same time, Figure~\ref{fig:xyplot_accuracy_pw} indicates that the accuracy of GLMM trees is unaffected by the number of covariates. Furthermore, GLMM tree performs best in recovering purely piecewise interactions, and poorest in recovering purely linear interactions. Finally, the performance of GLMM tree is affected by sample size: generally, the true effects are more often recovered in datasets with larger sample sizes.

@@

Both algorithms showed equal average correlations between the true and predicted treatment-effect differences: average accuracy was .60 (SD = .36) for GLMMs and .60 (SD = .32) for GLMM trees. In terms of bias, GLMM treatment-effect difference predictions deviated more from the true values than the predictions of GLMM tree. The mean true treatment-effect difference averaged over all datasets was .04 (SD = 1.22), whereas the mean predicted treatment-effect differences averaged over all datasets for GLMMs were -.26 (SD = 100.33) and for GLMM trees .03 (SD = 3.88). Furthermore, it should be noted that the models used for prediction with GLMM trees were more complex than the GLMMs used for prediction. Predictions for GLMMs in datasets with 10 covariates involved 11 (significant or non-significant) main effects, 100 first-order interactions and 100 second-order interactions. As the maximum tree depth of GLMM trees was set to 4, GLMM trees used a maximum of only 8 terminal nodes for predictions.

The most important predictors of predictive accuracy were algorithm type, sample size, the number of covariates, and type of interaction (linear, piecewise or both). These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}. As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely linear interactions. Furthermore, performance of GLMM tree is hardly affected by the number of covariates, whereas the predictive accuracy of GLMMs deteriorates when the number of covariates increases. This indicates that GLMM tree may be especially useful for exploratory purposes.

>>>>

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are particularly useful for exploratory purposes, as they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions. Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov15, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.

In many instances, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. For tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection allow for taking into account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM tree.

The GLMM tree algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}), which offers a flexible framework for subgroup detection. For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}. In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14), GLMM tree allows for taking into account the clustered structure of datasets. In contrast to previously suggested regression trees with random effects, GLMM tree allows for treatment effect estimation, with continuous as well as non-continuous response variables \cite<e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we first introduce an artificial motivating dataset, which will be used to illustrate and explain treatment-subgroup interaction detection with GLM-based MOB and GLMM tree. In the Simulation-based Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects. The comparison with GLM-based MOB allows for assessing the extent to which including random-effects estimation improves the performance of MOB trees. The comparison with GLMMs with pre-specified interactions will allow for assessing the extent to which using a tree-based method will improve detection of treatment-subgroup interactions. In the Application, we apply the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. Finally, in the Discussion we summarize the results and point out some directions for future research.

@@

Participants were part of one of ten clusters, each with a different value for the random intercept. Data were generated such that covariates and cluster-specific intercepts were uncorrelated. 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, while 8\% of variance was due to cluster-specific variation.

@@

The rationale behind MOB is that a global parametric model may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. However, CART trees detect differences in the mean (or median) value across terminal nodes, whereas MOB trees detect differences in parameters of more complex models across terminal nodes.

For example, let us take a global GLM to estimate the overall treatment effect in the motivating dataset. The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function\footnote{An overview of notation used is provided in the appendix.}:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups. Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper. Also, we assume a continuous response variable, employing a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$. However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.

@@

The result of fitting a global GLM to the motivating dataset is depicted in Figure~\ref{fig:fixedeffects}; the boxplots show the distribution of the posttreatment depression scores in both treatment groups. The global model does not describe the data well: there is substantial residual variance and the slope of the regression line is nearly zero. This does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.

The MOB algorithm can be used to detect such moderation, by testing for parameter stability over a set of auxiliary covariates, also known as \textit{partitioning variables}. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

@@

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ yields a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). This partition is created using $U_{k*}$, the variable with the minimal p-value in step (2). The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node or subgroup:
%
\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}
%
Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all $x$-variables interact with the subgroup indicator. Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{*}$ would have length $2J$ as well, containing subgroup-specific fixed-effects coefficients.

@@

Figure~\ref{fig:example_mobtree} shows the GLM tree grown on the motivating dataset. By using the three auxiliary covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the true treatment-subgroup interactions described above. However, the two subgroups in the middle do not represent true subgroups, which may be due to the clustered structure of the dataset not being taken into account.

@@

For estimating GLMM trees, we take this approach a step further: instead of a CART tree with constant fits to estimate the fixed-effects part of the GLMM, we use a GLM tree. This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes. In addition, GLMM trees can be estimated for continuous, as well as binary and count variables. The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:

@@

In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented. As can be seen, by taking into account the clustering of observations, the true treatment subgroups have been recovered, and the spurious split involving the anxiety variable no longer appears in the tree.

@@

\section{Simulation-based Evaluation: Method}

To assess the performance of GLMM tree, we performed three simulation studies. In Study I, we compared the accuracy of GLMM tree with that of GLM tree in datasets with treatment-subgroup interactions. In Study II, we compared the accuracy of GLMM tree with that of GLM tree in datasets without treatment-subgroup interactions, allowing for assessment of the Type-I error rate. In Study III, we compared the accuracy of GLMM tree with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.

@@

R \cite{R14} was used for the generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions. For other response distributions, the \verb|glmtree| function would be available. For estimation of GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.

@@

\subsubsection{Continuous interactions}

The subgroups depicted in Figure~\ref{fig:modelC} are piecewise functions of the partitioning variables, therefore these interactions can be referred to as piecewise interactions. However, interactions may also consist of continuous functions of the partitioning variables. For comparing the accuracy of GLMM tree with that of GLMMs with pre-specified interaction terms, we generated datasets in which interactions were varied to be purely piecewise, purely continuous, or a combination of both. To generate purely continuous interactions, the outcome variable $y$ was generated as the sum of:

Note that the direction (positive or negative) and type (main or interaction) of these effects correspond to those in Figure~\ref{fig:modelC}. All interactions were created by using centered $U_k$ variables, calculated by subtracting the variable means.

@@

Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}) in each dataset. Consequently, predictive accuracy was only assessed in datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions. To prevent overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, predictive accuracy was assessed using test datasets. Test datasets were generated from the same population as training datasets, but test observations were not drawn from the same clusters as the training observations, but from `new' clusters.

\section{Simulation-based Evaluation: Results}

\subsection{Study I: Detection of Treatment-Subgroup Interactions}

For each cell of the data-generating design described above, 50 datasets were generated, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2$ = 32,400 training datasets with treatment-subgroup interactions. In these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure~\ref{fig:modelC}).

@@

Figure~\ref{fig:xyplot_treesize_interact} indicates that GLMM tree grows trees of about the true tree size in all conditions. On average, tree size increases slightly with sample size for GLMM tree, and dramatically for GLM tree. In the absence of random effects (i.e., $\sigma_{b}=0$), GLM and GLMM trees grow trees of about the same size. However, clear differences in tree size were observed when $\sigma_b > 0$. When sample size is small and random intercepts are not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average. When sample size is larger and random intercepts are not correlated with a partitioning covariate, GLM and GLMM trees are about equally sized. However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially with larger sample sizes (i.e., 500 or 1000) and larger $\sigma_b$ values (i.e., 10). This effect was somewhat stronger when the random intercept was correlated with a covariate that did not appear in the true tree.

@@

\subsubsection{Design}
In addition to the accuracy of treatment-subgroup interaction detection, we assessed the Type-I error: the extent to which datasets without treatment-subgroup interactions may erroneously be partitioned. For assessing Type-I error of GLM and GLMM tree, we generated datasets without treatment-subgroup interactions, in which there was only a main effect of treatment in the population. Put differently, there was only a single global value of $\beta_j = \beta$ in every dataset. For these datasets, we varied the same parameters of the data-generating process as in Study I. However, the sixth facet of the data-generating process had only two levels for these datasets, as there were no `true' partitioning variables in these datasets. Therefore, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2$ = 21,600 datasets without treatment-subgroup interactions were generated.

@@

\subsection{Study III: Piecewise vs. Continuous Interactions}

\subsubsection{Design}

To compare the performance of GLMM tree with that of GLMMs with pre-specified interaction effects, we created datasets in which interactions were varied from purely continuous to purely piecewise. For these datasets, the seventh facet of the interaction design (effect size of treatment differences) was replaced by a factor with three levels, controlling the type of interactions: purely continuous interactions, purely piecewise interactions, and a combination of both. Purely continuous interactions were generated as described above, in the Method section. To allow for a smooth transition from purely continuous to purely continuous interactions, we aimed to keep the proportion of variance explained by each of the covariates equal across piecewise, continuous and combined interactions. Therefore, for piecewise interactions, we used the partition depicted in Figure 4, but not the coefficients. The coefficients used for data-generation were replaced by the subgroup-specific means, given purely continuous interactions. To generate the combined conitnuous and piecewise interactions, $y$ values according to the purely piecewise and continuous interaction models were summed and divided by 2.

Furthermore, in this simulation study, the number of cells in the simulation design was reduced by limiting the fourth facet of the data-generating design to a single level ($M=25$ clusters), as the earlier simulations indicated no effects of the number of clusters. The fifth facet of the data-generating design was limited to two levels ($\sigma_b = 2.5$ and $\sigma_b = 7.5$). As in Study II, we limited the sixth facet to two levels (the random intercept was either correlated to one of the covariates, or not). To counter unidentifiability due to potentially large numbers of pre-specified interaction terms in the GLMMs, the second level of the second facet (number of potential partitioning covariates) was set to $k=10$, instead of the value of $k=15$ used in the earlier simulations. As a result, $50 \times 3 \times 2 \times 2 \times 1 \times 2 \times 2 \times 3$ = 7,200 training datasets were generated for this study.

@@

<<linear-simulation-data>>=
load("pw_treespecs_long.dat")
treespecs.long$N <- factor(as.numeric(as.character(treespecs.long$N)), ordered=T)
treespecs.long$sigmab <- factor(as.numeric(as.character(treespecs.long$sigmab)), ordered=T)
treespecs.long$type <- factor(treespecs.long$type, ordered = TRUE, levels = 
c("continuous", "both", "piecewise"))
@

@@

Overall, GLMM tree accurately recovered main and interaction effects in 53.58\% of datasets, wheareas GLMMs accurately recovered main and interaction effects in 34.81\% of datasets. To find the most important predictors of model accuracy, we performed an ANOVA, in which the algorithm type, sample size, the number of covariates and type of interaction (continuous, piecewise or both) were found to be most important. These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_accuracy_pw}.

Figure~\ref{fig:xyplot_accuracy_pw} shows a clear effect of interaction type: GLMM tree performs best in datasets with purely piecewise interactions, and poorest in datasets with purely continuous interactions. GLMMs perform best in datasets with purely continuous interactions, and poorest in datasets with purely piecewise interactions. Figure~\ref{fig:xyplot_accuracy_pw} also shows a clear effect of the number of covariates: with a large number of covariates, GLMMs never recover all true effects in the dataset, whereas they always recover the true effects when the number of covariates is small and interaction effects are at least partly continuous. At the same time, the accuracy of GLMM tree is unaffected by the number of covariates. Finally, the performance of both GLMM tree and GLMMs is affected by sample size: the effects are more often recovered in datasets with larger sample sizes.

@@

Both algorithms showed about equal average correlations between the true and predicted treatment-effect differences: average accuracy was .60 (SD = .36) for GLMMs and .60 (SD = .32) for GLMM trees. In terms of bias, GLMM treatment-effect difference predictions deviated more from the true values than the predictions of GLMM tree. The mean true treatment-effect difference averaged over all datasets was .04 (SD = 1.22), whereas the mean predicted treatment-effect differences averaged over all datasets for GLMMs were -.26 (SD = 100.33) and for GLMM trees .03 (SD = 3.88). Furthermore, it should be noted that the predictive models for GLMM trees were simpler than those of the GLMMs. Predictions for GLMMs in datasets with 10 covariates involved 11 (significant or non-significant) main effects, 100 first-order interactions and 100 second-order interactions. Predictions for GLMM tree involved a maximum  of only 8 terminal nodes, as their maximum depth was restricted.

The most important predictors of predictive accuracy were algorithm type, sample size, the number of covariates, and type of interaction (continuous, piecewise or both). These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}. As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely continuous interactions. Furthermore, performance of GLMM tree is hardly affected by the number of covariates, whereas the predictive accuracy of GLMMs deteriorates when the number of covariates increases. This indicates that GLMM tree may be especially useful for exploratory purposes.
